---
title: "May 2017 National Occupational Employment and Wage Estimates United States"
date: "7/3/2018"
output: pdf_document
fig_width: 15 
fig_height: 20 
---

```{r, message=FALSE, warning=FALSE}
library(scales)
library(readxl)
library(Amelia)
library(psych)
library(knitr)
library(ggplot2)
library(ggalluvial)

#loading file
M2017 <- read_excel('national_M2017_dl.xlsx')

#file structure
kable(describe(M2017))
```

Before the analysis, I categorized all the occupation using https://www.cityofmadison.com/dcr/documents/EEO-1JobCat.pdf to categorize the occupation into relevant category. 

```{r,message=FALSE, warning=FALSE}
#checking missing values
missmap(M2017)
```

Due to large amount of missing value in "HOURLY" and "ANNUAL" columns, it is suggested that to remove these two columns.

```{r, message=FALSE, warning=FALSE}
# using mean 
M2017_1 <- M2017[,c(2,3,4,5,7,8)]

#remove 1st row
M2017_1 <- M2017_1[-1,]


#convert into relevant data type
M2017_1$OCC_TITLE <- as.character(M2017_1$OCC_TITLE)
M2017_1$OCC_GROUP <- as.factor(M2017_1$OCC_GROUP)
M2017_1$CATEGORY <- as.factor(M2017_1$CATEGORY)
M2017_1$TOT_EMP <- as.numeric(M2017_1$TOT_EMP)
M2017_1$H_MEAN <- as.numeric(M2017_1$H_MEAN)
M2017_1$A_MEAN <- as.numeric(M2017_1$A_MEAN)

#remove missing value
M2017_1.na <- na.omit(M2017_1)

#total employment by group
ggplot(M2017_1.na, aes(x = OCC_GROUP, y = TOT_EMP, fill = OCC_GROUP)) + 
  geom_bar(stat = 'identity') +
  scale_y_continuous(limits = c(-0,30000000), labels = comma) +
  labs(title = "Total Employment by Occupation Group", x = 'Occupation Group', y = "Total Employment") +
  guides(fill = guide_legend(title = "Occupation Group"))
```


```{r,  message=FALSE, warning=FALSE}
#total employment by category
ggplot(M2017_1.na, aes(x = CATEGORY, y = TOT_EMP, fill = CATEGORY)) + 
  geom_bar(stat = 'identity') +
  scale_y_continuous(limits = c(-0,30000000), labels = comma) +
  labs(title = "Total Employment by Occupation Category", x = 'Occupation Category', y = "Total Employment") +
  guides(fill = guide_legend(title = "Occupation Category")) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```


```{r}
#hourly wage
summary(M2017_1.na$H_MEAN)

M2017_1.na$Hour_wage <- cut(M2017_1.na$H_MEAN,
                         breaks = c(-Inf, 10.21, 27.74, Inf),
                         labels = c("low", "medium", "high"))

#annual income
M2017_1.na$Income <- cut(M2017_1.na$A_MEAN,
                         breaks = c(-Inf, 49999, 99999, Inf),
                         labels = c("low", "medium", "high"))

```

Due to lack of information for the hourly wage in USA, I decided to use the mean instead. I categorized 10.21 as low hourly wages, > 27.74 <= as medium hourly wage, and anything above that is consider as high. To categorize the annual income, I used [Middle Class, Income Are You in the Middle Class?](https://www.thebalance.com/definition-of-middle-class-income-4126870) published by the website the balance https://www.thebalance.com/definition-of-middle-class-income-4126870. 

```{r}
#catgorize the data
df <- table(M2017_1.na$OCC_GROUP, M2017_1.na$Hour_wage)
df1 <- table(M2017_1.na$OCC_GROUP, M2017_1.na$Income)
df2 <- table(M2017_1.na$CATEGORY, M2017_1.na$Hour_wage)
df3 <-  table(M2017_1.na$CATEGORY, M2017_1.na$Income)

df <- as.data.frame(df)
df1 <- as.data.frame(df1)
df2 <- as.data.frame(df2)
df3 <- as.data.frame(df3)

colnames(df) <- c("Group", "Wage (per Hour)", "Freq")
colnames(df1) <- c("Category", "Income (Annual)", "Freq")
colnames(df2) <- c("Group", "Wage (per Hour)", "Freq")
colnames(df3) <- c("Category", "Income (Annual)", "Freq")

#plot
ggplot(df, aes(x = Group, y = Freq, fill = `Wage (per Hour)`)) +
  geom_bar(stat = "identity", position = 'dodge') +
  labs(title = "Occupation Group by Hourly wage", x = 'Occupation Group', y = "Frequency") 
```


```{r}
ggplot(df1, aes(x = Category, y = Freq, fill = `Income (Annual)`)) +
  geom_bar(stat = "identity", position = 'dodge') +
  labs(title = "Occupation Group by Annualy Income", x = 'Occupation Group', y = "Frequency") 
```


```{r}
ggplot(df2, aes(x = Group, y = Freq, fill = `Wage (per Hour)`)) +
  geom_bar(stat = "identity", position = 'dodge') +
  labs(title = "Occupation Category by Hourly Wage", x = 'Occupation Category', y = "Frequency") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
ggplot(df3, aes(x = Category, y = Freq, fill = `Income (Annual)`)) +
  geom_bar(stat = "identity", position = 'dodge') +
  labs(title = "Occupation Category by Annualy Income", x = 'Occupation Category', y = "Frequency") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r, fig.width=10, fig.height=8}
#sankey diagram for Wage per hour
df4 <- table(M2017_1.na$OCC_GROUP, M2017_1.na$CATEGORY, M2017_1.na$Hour_wage)
df4 <- as.data.frame(df4)
colnames(df4) <- c("Group", "Category", "Wage (per Hour)", "Freq")

ggplot(df4,
       aes(weight = Freq, axis1 = Group, axis2 = Category, axis3 = `Wage (per Hour)`)) +
  geom_alluvium(aes(fill = `Wage (per Hour)`), width = 1/12) +
  geom_stratum(width = 1/12, fill = "green", color = "black") +
  geom_label(stat = "stratum", label.strata = TRUE) +
  scale_x_continuous(breaks = 1:3, labels = c("Group", "Category", "Wage (per Hour)")) +
  scale_fill_brewer(type = "qual", palette = "Set1") +
  ggtitle("Wage Per Hour by Occupation May 2017 (USA)") +
  theme_bw()
```

```{r, fig.width=10, fig.height=8}
#sankey diagram for annual income
df5 <- table(M2017_1.na$OCC_GROUP, M2017_1.na$CATEGORY, M2017_1.na$Income)
df5 <- as.data.frame(df5)
colnames(df5) <- c("Group", "Category", "Income (Annual)", "Freq")

ggplot(df5,
       aes(weight = Freq, axis1 = Group, axis2 = Category, axis3 = `Income (Annual)`)) +
  geom_alluvium(aes(fill = `Income (Annual)`), width = 1/12) +
  geom_stratum(width = 1/12, fill = "green", color = "black") +
  geom_label(stat = "stratum", label.strata = TRUE) +
  scale_x_continuous(breaks = 1:3, labels = c("Group", "Category", "Income (Annual)")) +
  scale_fill_brewer(type = "qual", palette = "Set1") +
  ggtitle("Annual Income by Occupation May 2017 (USA)") +
  theme_bw()
```

